---
title: "spencerzhang_A6"
author: "Spencer Zhang"
date: "10/28/2021"
output: bookdown::html_document2
editor_options: 
  chunk_output_type: console
---

# Children's Lead Risk from housing - Inspecting and modeling

In this study PUMAs data is used to inspect households with high risk of lead contamination and create a model to predict household risk in order to provide guidance for future preventative resource distribution.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

```{r library}
library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r metadata,eval=F}
pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "YBL",
    "BLD",
    "TEN",
    "MV",
    "NP",
    "HINCP",
    "AGEP"
  )
)

saveRDS(pums_2019_1yr, "a6_pums.rds") #save to disk so dont have to run the previous chunk ever again
```

```{r metadata-loading}
pums_2019_1yr <-readRDS("a6_pums.rds")
```

```{r motify-data-for-sf}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F) #get CA puma data

sf_boundary <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME == "San Francisco") #get CA geographic info for mapping

sf_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[sf_boundary, ] %>% 
  st_drop_geometry() %>%
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf() #create sf object for CA

sf_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% sf_pumas$PUMACE10)
```

We focus on the San Francisco area (Fig. \@ref(fig:map-sf-pumas-1)).

```{r map-sf-pumas-1, fig.align='center',fig.cap="Map of studied area San Francisco"}
leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = sf_pumas
  ) %>%
  addMarkers(
    data = sf_pumas %>% 
      st_centroid()
  )
```

```{r map-sf-pumas-2, eval =F}
ggplot(sf_pumas) + geom_sf()
```

```{r}
sf_pums_clean <- sf_pums %>%
  filter(YBL %in% c("1","2","3")) %>% #filtering the years needed
  mutate(
    YBL = as.numeric(YBL),
    HINCP = as.numeric(HINCP),
    AGEP = as.numeric(AGEP)
         ) %>%
  group_by(SERIALNO, .drop = FALSE) %>%
  arrange(AGEP) %>% #order your df by AGEP
  summarise_all(first) %>%
  mutate(
    LR = case_when(
          AGEP < 6 & HINCP < 90000~1,
          TRUE ~ 0
    )
  )
```

```{r factoring}
sf_pums_clean <- sf_pums_clean %>% 
  mutate(
    BLD = as.factor(sf_pums_clean$BLD),
    TEN = as.factor(sf_pums_clean$TEN),
    MV = as.factor(sf_pums_clean$MV),
    PUMA = as.factor(sf_pums_clean$PUMA)
  )
```

A lead risk prediction model is built with factors building built year, type of housing, move-in date, and PUMA information to predict the lead risk of given household.

Summary of the model can be found below:

```{r}
pums_model <- glm(LR ~BLD+TEN+MV+PUMA, family = quasibinomial(), sf_pums_clean)
```

```{r}
summary(pums_model)
```

The coefficients of the model can be found in Table \@ref(tab:cof-table).

```{r cof-table}
cof2 <- summary(pums_model)$coefficients
knitr::kable(cof2,
             caption = "Cofficient Table",
             digits = 2)
```

```{r predict}
new_data = sample_n(sf_pums_clean, size = 1)
```

To test the model, we pick a random column from our dataset and feed it to our model. The predicted score of risk for the random data is `r predict(pums_model, new_data, type ="response")` , where its true value is `r new_data$LR` .

Now we feed our entire dataset into the model to check its Type1/Type2 errors (Table \@ref(tab:error-table)).

```{r}
sf_pums_clean <- sf_pums_clean %>%
  cbind(
    SCORE = predict(pums_model,sf_pums_clean, type = "response")
  )
```

```{r}
summary_2x2 <-
  sf_pums_clean %>% 
  mutate(
    LR = ifelse(
      LR == 1, 
      "Yes (Risk)", 
      "No (No Risk)"
    )
  ) %>% 
  pull(LR) %>% 
  table(sf_pums_clean$SCORE >= 0.1)
```

```{r error-table}
knitr::kable(summary_2x2,
             caption = "Type1/2 Table")
```

The households exceeding my threshold score are typically built in 1939 or earlier, living in one-family house (detached), occupied without payment of rent, and have been living in the house for 5-9 years.

The success rate is relatively low. Roughly 31 sets of postcards (not considering WGTP) will be mailed out based on the specified threshold. Only 13% of the postcards actually go to low-income households with children, and 11% of children at risk are likely to be reached.

To improve the strategy, we can lower the threshold value to improve type1 error in the expense of type2 error (which is relatively harmless).

```{r}
temp <- tempfile()
download.file("https://www2.census.gov/programs-surveys/acs/data/pums/2019/1-Year/csv_hca.zip",destfile = temp)

pums_hca_2019_1yr <- read_csv(unzip(temp,"psam_h06.csv"))

unlink(temp)
```


```{r}
test <- pums_hca_2019_1yr %>%
  group_by(SERIALNO) %>%
  select(starts_with("WGTP"))

sf_pums_clean <- sf_pums_clean %>%
  left_join(., test, by = "SERIALNO")
```

```{r}
model2 <- glm(LR ~BLD+TEN+MV+PUMA, family = quasibinomial(), sf_pums_clean)
```

```{r}
summary_rw <-
  sf_pums_clean %>% 
  select(LR, SCORE, starts_with("WGTP"), -WGTP.x, -WGTP.y) %>%
  group_by(LR, SC=SCORE>0.1) %>%
  summarise_all(sum)
```

