---
title: "2020 Census: Changes in population and racial makeup of Foster City neighborhoods"
author: "by Spencer Zhang"
date: "10/03/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
  body{
  font-family: Verdana;
  font-size: 12pt;
}
</style>

According to 2020 Decennial Census data, the Foster City population grew 10.7% from 2010 to 2020, with large increases in the #todo:race population.
The data lets us see these changes down to the census blocks — areas with about 12 to 145 residents, with a rare maximum of 1324 residents per block. Here’s how each of the roughly 266 Foster City census blocks have changed in the past decade. 
The population change in Foster City varies in both directions and amplitudes, with the largest increase of 59 people per acre and a rate of 323.7% in one block, and minor decrease rate in several others.
Data for some census blocks may have been impacted by the coronavirus pandemic, so we advise people to focus on overall trends rather than individual block-level details.
```{r, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F) #echo = F to mask the code chunks
```
```{r}
library(tigris)
library(tidyverse)
library(sf)
library(mapview)
library(leaflet)
library(censusapi)
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r}
#2020 metadata and pop data fetch from census
dec_vars_2020 <-
  listCensusMetadata(name = "2020/dec/pl",
    type = "variables")
smc_pop_2020 <-
  getCensus(name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P1_001N") %>% 
  transmute(
    GEOID20 = paste0(state,county,tract,block),
    pop_2020 = P1_001N
  )
blocks_2020 <- blocks("CA", "San Mateo", year = 2020, progress_bar = F)
```
```{r}
#2010 metadata and pop data fetch from census
dec_vars_2010 <-
  listCensusMetadata(name = "2010/dec/pl",
    type = "variables")
smc_pop_2010 <-
  getCensus(name = "dec/pl",
    vintage = 2010,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P001001") %>% 
  transmute(
    GEOID10 = paste0(state,county,tract,block),
    pop_2010 = P001001
  )
blocks_2010 <- blocks("CA", "San Mateo", year = 2010, progress_bar = F)
```

```{r}
#race data fetch from census
smc_pop_race_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "group(P1)"
  ) %>% 
  mutate(
    block =
      paste0(state,county,tract,block)
  ) %>% 
  select(!c(GEO_ID,state,county,tract,NAME) & !ends_with(c("NA"))) %>% 
  pivot_longer(
    ends_with("N"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    dec_vars_2020 %>% 
      select(name, label)
  ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,NA,"category1","category2"),
    sep = "!!"
  )
```
```{r}
smc_pop_race_2010 <-
  getCensus(
    name = "dec/pl",
    vintage = 2010,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "group(P1)"
  ) %>% 
  mutate(
    block =
      paste0(state,county,tract,block)
  ) %>% 
  select(!c(GEO_ID,state,county,tract,NAME) & !ends_with(c("NA"))) %>% 
  pivot_longer(
    starts_with("P"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    dec_vars_2010 %>% 
      select(name, label)
  ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,NA,"category1","category2"),
    sep = "!!"
  )
```

```{r}
#Foster City as the study object
#get foster city's boundary
fc_boundary <- places("CA", progress_bar = F) %>% 
  filter(NAME == "Foster City")

#get 2010 foster city's block geo info from smc
fc_blocks_2010 <- blocks_2010 %>% 
  st_centroid() %>% 
  .[fc_boundary, ] %>% 
  st_drop_geometry() %>% 
  left_join(blocks_2010 %>% select(GEOID10, geometry)) %>% 
  st_as_sf() %>% 
  select(GEOID10) %>% 
  left_join(smc_pop_2010 %>% select(GEOID10, pop_2010))

#mapview(fc_blocks_2010) + mapview(fc_blocks_2010, alpha.regions = 0, color = "red", lwd = 2)
```

```{r}
#get fc 2020 block&pop info
fc_blocks_2020 <- blocks_2020 %>%
  .[fc_boundary, ] %>%  #is nfo_boundary right?
  select(GEOID20) %>% 
  left_join(smc_pop_2020)
```

```{r}
#spatial subsetting
#cut 2020's block into 2010's boundary and sum the pop
  #basically matching 2020 pop into 2010 block
fc_blocks_2020_intersect <- fc_blocks_2020 %>% 
  st_transform(26910) %>% #transform to a planar coordinate so that we can cut the area up
  mutate(original_area = st_area(.)) %>% 
  st_intersection(
    fc_blocks_2010 %>%
      select(GEOID10) %>% 
      st_transform(26910)
  ) %>% #return the intersection geometry of b2020 and b2010 / cutting b2020 into more pieces using b2010 and join the 2010 columns to the return sf as well (here just GEOID10 cause we did 'select')
  mutate(
    leftover_area = st_area(.),
    perc_area = as.numeric(leftover_area / original_area),
    pop_2020 = pop_2020 * perc_area
  )
#return val
#names(fc_blocks_2020_intersect)
#[1] "GEOID20"       "pop_2020"      "original_area"
#[4] "GEOID10"       "geometry"      "leftover_area"
#[7] "perc_area"  
```

```{r}
#sum up the 2020 pop in the 2010 block
fc_blocks_2020_reshaped <- fc_blocks_2020_intersect %>%
  st_drop_geometry() %>% 
  group_by(GEOID10) %>% 
  summarize(
    pop_2020 = sum(pop_2020, na.rm=T) %>% round()
  ) 
#return val
#names(fc_blocks_2020_reshaped)
#[1] "GEOID10"  "pop_2020"
```

```{r}
#density change
projection <- "+proj=longlat +datum=WGS84 +units=ft +no_defs" #todo: dont know where this goes
fc_pop_2010_2020 <- fc_blocks_2010 %>% 
  left_join(fc_blocks_2020_reshaped) %>% 
  mutate(
    absolute_pop = (pop_2020 - pop_2010) / (as.numeric(st_area(.)) * 0.00024710538146717)
  )
#return
# "absolute_pop" in fc_pop_2010_2020
```

```{r}
#percentage change
#todo: change into pipeline
fc_pop_2010_2020$pop_perc = case_when( 
  fc_pop_2010_2020$pop_2010 == 0 & !fc_pop_2010_2020$pop_2020 == 0 ~100,
  fc_pop_2010_2020$pop_2010 == 0 & fc_pop_2010_2020$pop_2020 == 0 ~0,
  TRUE ~ (fc_pop_2010_2020$pop_2020 - fc_pop_2010_2020$pop_2010)/fc_pop_2010_2020$pop_2010
  )
```

```{r}
#Original comment
#To get our color pallete to be symmetrical across 0, we need the minimum and maximum bounds to be equidistant from 0. For our absolute change, let's try stretching one bound to meet the other. Using https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf, we can identify "PiYG" as a satisfactory palette for our purposes.

#get color pallete for density change
max <- max(abs(min(fc_pop_2010_2020$absolute_pop)), abs(max(fc_pop_2010_2020$absolute_pop)))
absolute_pal <- colorNumeric(
  palette = "PiYG",
  domain = c(-max,max)
)
```

## Population change in Foster City census blocks from 2010 to 2020
Mouse over a block for details.
```{r}
#map for density change
leaflet(data = fc_pop_2010_2020) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    fillColor = ~absolute_pal(absolute_pop),
    label = ~round(absolute_pop),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  ) %>% 
  addLegend(
    pal = absolute_pal,
    value = c(-max,max),
    title = "Population Density Change<br>in Foster City<br>from 2010 to 2020<br>(unit:per acre)")
```

## Population percentage change in Foster City census blocks from 2010 to 2020
Mouse over a block for details.
```{r}
perc_max <- max(abs(min(fc_pop_2010_2020$pop_perc)), abs(max(fc_pop_2010_2020$pop_perc)))
perc_pal <- colorNumeric(
  palette = "PiYG",
  domain = c(-perc_max,perc_max)
)

leaflet(data = fc_pop_2010_2020) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    fillColor = ~perc_pal(pop_perc),
    label = ~round(pop_perc),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  ) %>% 
  addLegend(
    pal = perc_pal,
    value = c(-perc_max,perc_max),
    title = "Population Percentage Change<br>in Foster City<br>from 2010 to 2020<br>(unit:%)")
```

```{r eval = F}
If you want to try multiple layers, you'll need to change some steps upstream to collect more than one field of information for each block. Then, you final leaflet step would look something like this:

#todo layering
leaflet(data = fc_pop_2010_2020) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    fillColor = ~absolute_pal(absolute_pop),
    label = ~round(absolute_pop),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  ) %>% 
  addPolygons(
    fillColor = ~perc_pal(pop_perc),
    label = ~round(pop_perc),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  ) %>% 
  addLegend(
    pal = absolute_pal,
    value = c(-max,max),
    title = "Population Density Change<br>in Foster City<br>from 2010 to 2020") %>% 
  addLayersControl(
    baseGroups = c("___","___"),
    position = "___",
    options = layersControlOptions(
      collapsed = F
    )
  )

You can also put layers in groups. But note: `addLayersControl()` currently does not support the toggling on/off of layers when you use `baseGroups`, which are the radio button selections. Instead you'd need to use `overlayGroups`, which are checkboxes.
Collapse
```
