---
title: "Difference in Differences analysis: Antioch BART station"
author: "Spencer Zhang"
date: "11/11/2021"
output: bookdown::html_document2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

# Background
Means of transportation plays a big part in daily commuters' life as well as in urban systems. In this report we look deeply into public transportation accessibility and its utilization. How much imapct does a transportation system have on commuter behavior? We take the newly built Antioch BART station as an example, and see the difference in the number of BART commuters before and after it's built (mid-2018).
```{r prep}
library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```

The area of interest is marked in blue shade in the map below. The newly built BART station location is marked by the blue pointer.
```{r area-map}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.78008514465182, #change this to your specific point(you can get the lng/lat data from google map)
    lat = 37.99574468958323
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )
```

```{r pums-data}
pums_2014_2019 <- readRDS("pums_2014_2019_wts.rds")
pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )
```

The map below shows the number of population in 2017 in each census tract in the Bay Area.
```{r pop-map}
pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```

```{r}
#creating your own sf object with geo data (rename the col names for sf to recognize)
#dataframe %>% st_as_sf(coords = c("Ing_name", "lat_name"), crs = 4326) 
```

The map below shows the number of BART takers in 2017 in each census tract in the Bay Area. Darker shade indicates more BART commuter. As can infer from the map, our tract of interest does not show heavy BART commute in year 2017.
```{r bart-map}
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2017) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2017) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )
```

```{r}
pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(
    names_from = year,
    values_from = bart
  )
```

```{r}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`,`2017`) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% #arrange by closest to most distance
  .[1:11, ] %>% #take the ten closest neighbor
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% #get geographic data for graphing
  st_as_sf() #change into a sf object for leaflet graphing
```

Intuitively, the number of population and BART commuters changes as time progresses, mostly in increasing manners. In order to separate the impact time has on the number of BART takers from the causal impact the Antioch BART station has on it, we chose ten Census tract with most similar time progress patterns as control group, using their number of BART takers after 2018 as base level to estimate how much the BART taker number would change if there was not a new station built.
The similar tracts chosen are shown below in blue, while red is our tract of interest.
```{r matching-pair-map}
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    #color = "red", #this line wasnt originally in the textbook code
    label = ~PUMA
  )
```

The change of number of BART takers from year 2014 to 2019 is shown below. The dotted line marks the year the Antioch BART station is built. Hypothetically, the number of BART takers in Antioch would progress from 2017 to 2018 in similar patterns as its control neighborhoods, as shown in blue. However, the actual number change differs vastly from its neighbors. The difference between the red line and the blue line may be attribute to the effect of the new Antioch BART station, which increases the BART accessibility in Antioch thus motivates more population in this given tract to take BART.
```{r}
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") +
  labs(
    title = "Antioch vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```

The coefficients of the regression model formula(BART ~ treated*time) is given below, where treated indicates new BART station being built, time indicates the effect time has on population increase and the corresponding BART taker number increase.
```{r model-1}
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year", #column names bacome a new column variable names year
    values_to = "bart" #the values used to correspond to each columns are now stored in bart column
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2017, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```
As compared to the example in textbook, the Antioch BART station has a much bigger impact on its neighborhood BART takers than the Newark station has on Milpitas, and the impact time and BART station has on the increase is more significant than the textbook as well. The number of BART takers in Antioch was much higher than Milpitas before the stations were built, indicating a possible larger population of commuters. The need for more accessible BART and other public transportation resources may be higher in Antioch than Milpitas.

# Considering replicate weights
The census data provided us with replicate weights to help with uncertainty and confidence interval. Below we repeat the previous regression analysis with replicate weights in consideration.
```{r}
replicate <- pums_2014_2019 %>%
  filter(JWTR %in% c("4"), PUMA %in% match$PUMA) %>%
  select(PUMA, year, starts_with("PW")) %>%
  group_by(PUMA, year) %>% 
  summarise_all(as.numeric) %>%
  group_by(PUMA, year) %>% 
  summarise_all(sum)
```

```{r}
transit_did_2 <-
  transit_did %>%
  left_join(., replicate, by = c("PUMA", "year"))
```

## method1
```{r}
transit_did_2[is.na(transit_did_2)] = 0

design <- svrepdesign(
  data = transit_did_2,
  type = "ACS",
  repweight = transit_did_2[, 7:86],
  weights = ~PWGTP
)
```

```{r}
model <- svyglm(
  formula = bart ~time*treated,
  family = gaussian(), #normal distribution
  design = design
)
```

The coefficients of the model are shown below. Under the model's prediction, the number of BART taker would increase by 688 after a year of a newly built BART station. The number of BART taker increase caused solely by new station would be 351.
The corrected standard errors are shown in the table below.
```{r}
summary(model)
```

## method2
This is the same analysis under a different coding progress.
```{r}
replicates_model <-
  1:80 %>% 
  map_dfr(function(x){
    transit_did_2 %>% 
      rename(replicate = paste0("PWGTP", x)) %>% 
      lm(
        formula = bart ~ treated*time,
        data = .,
        weights = replicate
      ) %>%
      .$coefficients %>%
      as.data.frame() %>%
      rownames_to_column()
  })
```

```{r}
weighted_model <- transit_did_2 %>%
  lm(
    formula = bart ~ treated*time,
    data = .,
    weights = as.numeric(PWGTP)
  )
```

```{r}
weighted_model_summary <- replicates_model %>%
  rename(replicate = ".") %>%
  left_join(
    weighted_model$coefficients %>%
      as.data.frame() %>%
      rownames_to_column() %>%
      rename(base = ".")
  ) %>%
  group_by(rowname) %>%
  summarize(
    stderr = sqrt(4/80*sum((base - replicate)^2)),
    base = first(base)
  ) %>%
  select(rowname, base, stderr)
```

```{r}
weighted_model <- transit_did_2 %>%
  lm(
    formula = bart ~ treated*time,
    data = .,
    weights = as.numeric(PWGTP)
  )
```

```{r}
knitr::kable(weighted_model_summary)
```

